Supp mat: time series paper

### 2017-08-19 writing/supp_mat_draft.Rmd
Final version of manuscript for PeerJ preprint

\begin{center}
 \textbf{Supplementary Material: Article title here}

@@

Data and code for all the figures and tables can be found at (https://github.com/erwhite1).

@@

Here, we illustrate how we performed the subsampling and power calculations for a specific population. We provide more detail here than in the main manuscript. As an example, we examine a 35-year time series of Bigeye tuna (\emph{Thunnus obesus}), one species in the Global Population Dynamics Database [@GPDD2010]. <!--Simple linear regression indicates a significant increase for this population with an estimated intercept of `r as.character(round(as.numeric(coef(model_example)[1])))` and a slope coefficient of `r as.numeric(coef(model_example)[2])`.--> We assume that this significant increase over 35 years is in fact the ``true trend". In statistical jargon, the 35-year trend is an effect that is actually present, we can reject the null hypothesis of no trend. We can then use this as a benchmark to see if subsamples of the time series also indicate a significant increase.

We first extract all contiguous subsamples of the time series. This leads to 34 two-year subsamples, 33 two-year subsamples, and so forth until a single 35-year subsample. We can call each set of subsamples, of the same length, a set. For each subsample, we conduct linear regression and extract model coefficients and p-values. Then, the fraction of subsamples within a set that show significant trends (significant slope coefficient) is the statistical power. It is important to note that we only consider subsamples to be significant if they are significant in the same direction as the complete 35-year time series. In other words, we are conducting a one-tailed test.

We can then plot statistical power as a function of time series length (Fig. \ref{fig:empirical_approach_example}b). As expected, we can see that power increases with the more years that are sampled. Then, we determine an appropriate level of statistical power that we find acceptable. Traditionally, this has been at 0.8, however, this is purely historical. Statistical power of 0.8 implies if a true trend is present, or a real change in abundance, then we will detect this trend 0.8 fraction of the time.

@@

With statistical power of 0.8, we then determine the minimum time series length ($T_{min}$) required to achieve that level of statistical power. Here, $T_{min}$ is the first point in Fig. \ref{fig:empirical_approach_example}b where following points are also above 0.8. In this example, $T_{min}$ is `r head(which(power_example>0.8),1)`. Therefore, a minimum of 22 years of continuous monitoring are required to have 0.8 statistical power in determining long-term changes in abundance.

@@

## Statistical model of minimum time required and time series characteristics

In the main text, we explained how the minimum time required strongly correlated with the trend strength, temporal autocorrelation, and variance in population size. Here, we use a generalized linear model framework with a Poisson error structure to determine drivers of the minimum time required. In figure \ref{fig:poisson_model}, we show a set of residual plots for the regression. We then show the coefficient estimates and levels of significance in table \ref{table:model_output}.

@@

@@

In the main manuscript, we examined the minimum time required to detect a significant trend in abundance over time using linear regression. As detailed in the main manuscript the minimum time required was around 15, but there was a wide distribution. Therefore, we were interested in potential explanatory variables of the minimum time required. In the main manuscript, we examined characteristics of the time series itself, like variability, autocorrelation, and the trend in abundance over time. Here, we combined our time series database with a database on life history characteristics of amniotes from @Myhrvold2015. There was life history information available for `r nrow(LPI_pop_info)` populations representing `r length(table(LPI_pop_info$Binomial))` different species, all of which were in the Aves class.

We then correlated minimum time required for each population with its corresponding life history characteristics. In figure \ref{fig:biological_correlates} we examined minimum time required versus generation length (years), litter size (n), adult body mass (grams), maximum longevity (years), egg mass (grams), and incubation (days). None of these variables had much explanatory power in accounting for the variance in the minimum time required. see table \ref{table:model_output}}

@@

@@

When we examined differences in the minimum time required there were similar patterns with biological classes. For instance, more time is required for species within the Actinopterygii class compared to other species (Fig. \ref{fig:class}a). These differences between biological classes can be explained by differences in population variability, with species in Actinopterygii have larger variability in population size from year-to-year.

@@

load("../cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
# Biological correlates plots
par(mfrow=c(2,2),oma=c(8,1,1.5,0.5),mar=c(0,4,0,1))

@@

Estimates of $T_{min}$ depend strongly on the values used for the significance level ($\alpha$) and the probability of type II error ($\beta$), both of which are set by the practitioner. Here we explore how estimates of $T_{min}$ are affected by changes in each of these parameters. We see that the minimum time required increases with increases in statistical power or decreases with increases in the threshold for statistical significance (Fig \ref{fig:min_time_vs_alpha_beta}).

@@

In the main text we evaluated the minimum time required to determine long-term trends in abundance via linear regression. This process examined linear trends in abundance over time. Here we examine the minimum time required to estimate long-term trends that are either exponential growth or decay. We use the same methods as described in the previous sections, but we take the $log$ of population density, or abundance.

@@

We see that the distribution of $T_{min}$ is almost identical to that in the main manuscript. This is perhaps not surprising as most time series that would significantly increase or decrease linearly would probably also significantly increase or decrease at an exponential rate. Further, the calculations here and in the main manuscript both use linear regression, and both estimate the same number of parameters.

@@

In the main text, we showed how a simple population model could be simulated repeatedly to estimate the power obtained with time series of increasing length. The model in the main text simulated linear population growth with only a slope coefficient, y-intercept, and noise parameter required. This model is purely phenomenological and does not include any species life history. Here, we use the same routine as the main text, but simulate from a more biologically-realistic population model. We simulate the model described in @White2014. The model is a stochastic, age-structured population model that includes density-dependence for lemon sharks in Bimini, Bahamas.

@@

Here, we parameterize the model for a situation where adult mortality rate is high enough to cause a population decline. In the same way we simulated models in the main text, we simulate this more biologically-realistic model for different lengths of time. For each length of time, we calculate the statistical power. Similar to results with the simpler model, statistical power generally increases with longer sampling time (Fig. \ref{fig:shark_example}). In this example, the minimum time required ($T_{min}$) to obtain at least 0.8 statistical power, given a significance level of 0.05, is `r years_to_sample[tail(which(shark_power<0.8),1)+1]` years.

@@

We applied the same calculations to `r nrow(pop_info)` populations, as we did in the main text. We then obtain a distribution of the minimum time required to measure the ``true" long-term growth rate (Fig. \ref{fig:min_time_growth_dist}). We see a bimodal distribution with many populations required 30+ years to estimate the long-term growth rate. The large number of short years required is due to cases where the entire time series is consistently increasing or decreasing at the same rate.

@@

In the main text, we examined the minimum time required to identify a trend in abundance via linear regression. This approach allowed us to identify increases or decreases, but a linear model may not always be a good fit. Generalized additive models (GAMs) are more general than general linear models and allow more flexible. GAMs are models where a response variable depends on unknown smooth functions of explanatory variables. GAMs, therefore, can identify relationships between response variables and explanatory variables that are non-linear and perhaps more complicated. The downside of GAMs is they typically require more data and are also prone to overfitting (citations).

Here, we conduct the same analyses in the main text, but instead calculate the minimum time series required to detect trends over time according to a GAM model. We hypothesize that GAMs should require less time to detect a trend as they are more flexible than linear regression. We provide an example that shows statistical power increases with more time sampled (Fig. \ref{fig:gam_example}).

>>>>

\begin{center}
 \textbf{Supplementary Material: Minimum time required to detect population trends: the need for long-term monitoring programs}

@@

Data and code for all the figures and tables can be found at (https://github.com/erwhite1/time-series-project). All analyses were run using R [@RCoreTeam2016].

@@

Here, we illustrate how we performed the subsampling and power calculations for a specific population. We provide more detail here than in the main manuscript. As an example, we examine a 35-year time series of Bigeye tuna (\emph{Thunnus obesus}), one species in the Global Population Dynamics Database [@GPDD2010]. Simple linear regression indicates a significant decrease for this population with an estimated slope coefficient of `r as.numeric(coef(model_example)[2])`. We assume that this significant increase over 35 years is in fact the ``true trend". In statistical jargon, the 35-year trend is an effect that is actually present; we can reject the null hypothesis of no trend. We can then use this as a benchmark to see if subsamples of the time series also indicate a significant increase.

@@

We then perform a subsampling routine to estimate the minimum time required $T_{min}$ (similar to @Gerber1999). This is the same routine we used for the results in the main manuscript.
\begin{enumerate}
  \item We first extract all contiguous subsamples of the time series. This leads to 34 two-year subsamples, 33 two-year subsamples, and so forth until a single 35-year subsample.  
  \item For each subsample, we conduct linear regression and extract model coefficients and p-values.    \item We can call each set of subsamples, of the same length, a set. The fraction of subsamples within a set that show significant trends (significant slope coefficient $\alpha$ less than 0.05) is the statistical power. It is important to note that we only consider subsamples to be significant if they are significant in the same direction as the complete 35-year time series. In other words, we are conducting a one-tailed test.
  \item We can then plot statistical power as a function of time series length (Fig. \ref{fig:empirical_approach_example}b). As expected, we can see that power increases with the more years that are sampled.
  \item  Then, we determine an appropriate level of statistical power that we find acceptable. Traditionally, this has been at 0.8, however, this is purely historical. Statistical power of 0.8 implies if a true trend is present, or there is a real change in abundance, then we will detect this trend 0.8 fraction of the time. 
  \item We then determine the minimum time series length ($T_{min}$) required to achieve that level of statistical power. Here, $T_{min}$ is the first point in Fig. \ref{fig:empirical_approach_example}b where following points are also above 0.8. In this example, $T_{min}$ is `r head(which(power_example>0.8),1)`. 
\end{enumerate}

Therefore, a minimum of `r head(which(power_example>0.8),1)` years of continuous monitoring are required (for 0.8 statistical power at 0.05 significance level) to determine long-term changes in abundance.

@@

## Predictors of minimum time required

@@

In the main text, we explained how the minimum time required strongly correlated with the trend strength, temporal autocorrelation, and variance in population size. Here, we use a generalized linear model framework with a Poisson error structure to determine explanatory variables of the minimum time required. We use the same `r nrow(pop_info)` populations as in the main text. In figure \ref{fig:poisson_model}, we show a set of residual plots for the regression. We then show the coefficient estimates and levels of significance in table \ref{table:model_output}.

@@

In the main manuscript, we examined the minimum time required to detect a significant trend in abundance over time using linear regression. As detailed in the main manuscript the minimum time required was around 15, but there was a wide distribution. Therefore, we were interested in potential explanatory variables of the minimum time required. In the main manuscript, we examined characteristics of the time series itself, like variability, autocorrelation, and the trend in abundance over time. Here, we combined our time series data with a data on life history characteristics of amniotes from @Myhrvold2015. There was life history information available for `r nrow(LPI_pop_info)` populations representing `r length(table(LPI_pop_info$Binomial))` different species, all of which were birds (Aves class).

@@

#######

poisson_model <-glm(min_time_for_power~ gen_len + litter_or_clutch_size_n + adult_body_mass_g +maximum_longevity_y + egg_mass_g + incubation_d,family = 'poisson',data=LPI_pop_info)

#  par(mfrow=c(2,2),mar=c(5,5,1,2),oma=c(0,0,0.5,0))
#    plot(poisson_model,cex.lab=1.2,las=1,cex.axis=1.2,main=' ',pch=16, col=rgb(0.5,0.5,0.5,0.4))
#R^2 calculation (pg. 218 Zuur et al. 2009 mixed effects book)
R2_fullmodel <- 100*(poisson_model$null.deviance - poisson_model$deviance)/(poisson_model$null.deviance)
```

We then correlated minimum time required for each population with its corresponding life history characteristics. In figure \ref{fig:biological_correlates} we examined minimum time required versus generation length (years), litter size (n), adult body mass (grams), maximum longevity (years), egg mass (grams), and incubation (days). None of these variables had much explanatory power in accounting for the variance in the minimum time required. We ran a generalized linear model (with a Poisson error structure) and found that these six variables only accounted for `r round(100*(poisson_model$null.deviance - poisson_model$deviance)/(poisson_model$null.deviance),2)`% of the explained deviance in minimum time required.

@@

Minimum time required differed for different biological classes. For instance, more time is required for species within the Actinopterygii class compared to other species (Fig. \ref{fig:class}a). These differences between biological classes can be explained by differences in population variability, with species in the Actinopterygii class experiencing larger inter-annual variability in population size (Fig. \ref{fig:class}c).

@@

load("../cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
# Biological correlates plots
par(mfrow=c(2,2),oma=c(8,1,1.5,0.5),mar=c(0.5,4,0,1))

@@

Estimates of $T_{min}$ depend strongly on the values used for the statistical significance level ($\alpha$) and the probability of type II error ($\beta$), both of which are set by the practitioner. Again, statistical power is $1-\beta$. Here we used simulations of the model described in the main manuscript. The model simulates linear trends in population abundance. We explored how estimates of $T_{min}$ are affected by changes in each of these parameters. We see that the minimum time required increases with increases in statistical power or decreases with increases in the threshold for statistical significance (Fig \ref{fig:min_time_vs_alpha_beta}).

@@

In the main text, we evaluated the minimum time required to determine long-term trends in abundance via linear regression. This process examined linear trends in abundance over time. Here we examine the minimum time required to estimate long-term trends that are either exponential growth or decay. We use the same methods as described in the previous sections, but we take the $log$ of population density, or abundance.

@@

We see that the distribution of $T_{min}$ is almost identical to that in the main manuscript (Fig. \ref{fig:min_time_dist_log_pop}). This is perhaps not surprising as most time series that would significantly increase or decrease linearly would probably also significantly increase or decrease at an exponential rate. Further, the calculations here and in the main manuscript both use linear regression. Therefore both calculations estimate the same number of parameters.

@@

In the main text, we showed how a simple population model could be simulated repeatedly to estimate the power obtained with time series of increasing length [@Bolker2008;@Johnson2015]. The model in the main text simulated linear population growth with only a slope coefficient, y-intercept, and noise parameter required. This model is purely phenomenological and does not include any species life history. Here, we use the same routine as the main text, but simulate from a more biologically-realistic population model. We use the model described in @White2014. The model is a stochastic, age-structured population model that includes density-dependence for lemon sharks (\emph{Negaprion brevirostris}) in Bimini, Bahamas.

@@

Here, we parameterize the model for a situation where adult mortality rate is high enough to cause a population decline. In the same way we simulated models in the main text, we simulate this more biologically-realistic model for different lengths of time. For each length of time, we calculate the statistical power. Similar to results with the simpler model, statistical power generally increases with longer sampling time (Fig. \ref{fig:shark_example}). In this example, the minimum time required ($T_{min}$) t0 obtain at least 0.8 statistical power, given a significance level of 0.05, is `r years_to_sample[tail(which(shark_power<0.8),1)+1]` years.

@@

We applied the same calculations to `r nrow(pop_info)` population time series. We then obtain a distribution of the minimum time required to measure the ``true" long-term growth rate (Fig. \ref{fig:min_time_growth_dist}). We see a bimodal distribution with many populations required 30+ years to estimate the long-term growth rate. The large number of short years required is due to cases where the entire time series is consistently increasing or decreasing at the same rate.

@@

In the main text, we examined the minimum time required to identify a trend in abundance via linear regression. This approach allowed us to identify increases or decreases, but a linear model may not always be a good fit. Generalized additive models (GAMs) are more general than general linear models and allow more flexibility [@Wood2006]. GAMs are models where a response variable depends on unknown smooth functions of explanatory variables. GAMs, therefore, can identify relationships between response variables and explanatory variables that are non-linear and perhaps more complicated. The downside of GAMs is they typically require more data and are also prone to overfitting.

Here, we conduct the same analyses in the main text, but instead calculate the minimum time series required to detect trends over time according to a GAM model. We hypothesized that GAMs should require less time to detect a trend as they are more flexible than linear regression. We provide an example that shows statistical power increases with more time sampled (Fig. \ref{fig:gam_example}).

### 2017-08-19 writing/supp_mat_draft.tex
Final version of manuscript for PeerJ preprint

Data and code for all the figures and tables can be found at
(\url{https://github.com/erwhite1}).

@@

Here, we illustrate how we performed the subsampling and power
calculations for a specific population. We provide more detail here than
in the main manuscript. As an example, we examine a 35-year time series
of Bigeye tuna (\emph{Thunnus obesus}), one species in the Global
Population Dynamics Database (NERC Centre for Population Biology 2010).
We assume that this significant increase over 35 years is in fact the
``true trend``. In statistical jargon, the 35-year trend is an effect
that is actually present, we can reject the null hypothesis of no trend.
We can then use this as a benchmark to see if subsamples of the time
series also indicate a significant increase.

We first extract all contiguous subsamples of the time series. This
leads to 34 two-year subsamples, 33 two-year subsamples, and so forth
until a single 35-year subsample. We can call each set of subsamples, of
the same length, a set. For each subsample, we conduct linear regression
and extract model coefficients and p-values. Then, the fraction of
subsamples within a set that show significant trends (significant slope
coefficient) is the statistical power. It is important to note that we
only consider subsamples to be significant if they are significant in
the same direction as the complete 35-year time series. In other words,
we are conducting a one-tailed test.

We can then plot statistical power as a function of time series length
(Fig. \ref{fig:empirical_approach_example}b). As expected, we can see
that power increases with the more years that are sampled. Then, we
determine an appropriate level of statistical power that we find
acceptable. Traditionally, this has been at 0.8, however, this is purely
historical. Statistical power of 0.8 implies if a true trend is present,
or a real change in abundance, then we will detect this trend 0.8
fraction of the time.

@@

With statistical power of 0.8, we then determine the minimum time series
length (\(T_{min}\)) required to achieve that level of statistical
power. Here, \(T_{min}\) is the first point in Fig.
\ref{fig:empirical_approach_example}b where following points are also
above 0.8. In this example, \(T_{min}\) is 22. Therefore, a minimum of
22 years of continuous monitoring are required to have 0.8 statistical
power in determining long-term changes in abundance.

@@

\subsection{Statistical model of minimum time required and time series
characteristics}\label{statistical-model-of-minimum-time-required-and-time-series-characteristics}

In the main text, we explained how the minimum time required strongly
correlated with the trend strength, temporal autocorrelation, and
variance in population size. Here, we use a generalized linear model
framework with a Poisson error structure to determine drivers of the
minimum time required. In figure \ref{fig:poisson_model}, we show a set
of residual plots for the regression. We then show the coefficient
estimates and levels of significance in table \ref{table:model_output}.

@@

@@

In the main manuscript, we examined the minimum time required to detect
a significant trend in abundance over time using linear regression. As
detailed in the main manuscript the minimum time required was around 15,
but there was a wide distribution. Therefore, we were interested in
potential explanatory variables of the minimum time required. In the
main manuscript, we examined characteristics of the time series itself,
like variability, autocorrelation, and the trend in abundance over time.
Here, we combined our time series database with a database on life
history characteristics of amniotes from Myhrvold et al. (2015). There
was life history information available for 547 populations representing
315 different species, all of which were in the Aves class.

We then correlated minimum time required for each population with its
corresponding life history characteristics. In figure
\ref{fig:biological_correlates} we examined minimum time required versus
generation length (years), litter size (n), adult body mass (grams),
maximum longevity (years), egg mass (grams), and incubation (days). None
of these variables had much explanatory power in accounting for the
variance in the minimum time required. see table
\ref{table:model_output}\}

@@

@@

Estimates of \(T_{min}\) depend strongly on the values used for the
significance level (\(\alpha\)) and the probability of type II error
(\(\beta\)), both of which are set by the practitioner. Here we explore
how estimates of \(T_{min}\) are affected by changes in each of these
parameters. We see that the minimum time required increases with
increases in statistical power or decreases with increases in the
threshold for statistical significance (Fig
\ref{fig:min_time_vs_alpha_beta}).

@@

In the main text we evaluated the minimum time required to determine
long-term trends in abundance via linear regression. This process
examined linear trends in abundance over time. Here we examine the
minimum time required to estimate long-term trends that are either
exponential growth or decay. We use the same methods as described in the
previous sections, but we take the \(log\) of population density, or
abundance.

@@

We see that the distribution of \(T_{min}\) is almost identical to that
in the main manuscript. This is perhaps not surprising as most time
series that would significantly increase or decrease linearly would
probably also significantly increase or decrease at an exponential rate.
Further, the calculations here and in the main manuscript both use
linear regression, and both estimate the same number of parameters.

@@

In the main text, we showed how a simple population model could be
simulated repeatedly to estimate the power obtained with time series of
increasing length. The model in the main text simulated linear
population growth with only a slope coefficient, y-intercept, and noise
parameter required. This model is purely phenomenological and does not
include any species life history. Here, we use the same routine as the
main text, but simulate from a more biologically-realistic population
model. We simulate the model described in White, Nagy, and Gruber
(2014). The model is a stochastic, age-structured population model that
includes density-dependence for lemon sharks in Bimini, Bahamas.

@@

Here, we parameterize the model for a situation where adult mortality
rate is high enough to cause a population decline. In the same way we
simulated models in the main text, we simulate this more
biologically-realistic model for different lengths of time. For each
length of time, we calculate the statistical power. Similar to results
with the simpler model, statistical power generally increases with
longer sampling time (Fig. \ref{fig:shark_example}). In this example,
the minimum time required (\(T_{min}\)) to obtain at least 0.8
statistical power, given a significance level of 0.05, is 27 years.

@@

We applied the same calculations to 1032 populations, as we did in the
main text. We then obtain a distribution of the minimum time required to
measure the ``true" long-term growth rate (Fig.
\ref{fig:min_time_growth_dist}). We see a bimodal distribution with many
populations required 30+ years to estimate the long-term growth rate.
The large number of short years required is due to cases where the
entire time series is consistently increasing or decreasing at the same
rate.

@@

In the main text, we examined the minimum time required to identify a
trend in abundance via linear regression. This approach allowed us to
identify increases or decreases, but a linear model may not always be a
good fit. Generalized additive models (GAMs) are more general than
general linear models and allow more flexible. GAMs are models where a
response variable depends on unknown smooth functions of explanatory
variables. GAMs, therefore, can identify relationships between response
variables and explanatory variables that are non-linear and perhaps more
complicated. The downside of GAMs is they typically require more data
and are also prone to overfitting (citations).

Here, we conduct the same analyses in the main text, but instead
calculate the minimum time series required to detect trends over time
according to a GAM model. We hypothesize that GAMs should require less
time to detect a trend as they are more flexible than linear regression.
We provide an example that shows statistical power increases with more
time sampled (Fig. \ref{fig:gam_example}).

@@

\hypertarget{refs}{}
\hypertarget{ref-Myhrvold2015}{}
Myhrvold, Nathan P., Elita Baldridge, Benjamin Chan, Dhileep Sivam,
Daniel L. Freeman, and S.K. Morgan Ernest. 2015. ``An amniote
life-history database to perform comparative analyses with birds,
mammals, and reptiles.'' \emph{Ecology} 96 (11): 3109.

@@

@@

>>>>

Data and code for all the figures and tables can be found at
(\url{https://github.com/erwhite1/time-series-project}). All analyses
were run using R (R Core Team 2016).

@@

Here, we illustrate how we performed the subsampling and power
calculations for a specific population. We provide more detail here than
in the main manuscript. As an example, we examine a 35-year time series
of Bigeye tuna (\emph{Thunnus obesus}), one species in the Global
Population Dynamics Database (NERC Centre for Population Biology 2010).
Simple linear regression indicates a significant decrease for this
population with an estimated slope coefficient of -0.0189581. We assume
that this significant increase over 35 years is in fact the ``true
trend``. In statistical jargon, the 35-year trend is an effect that is
actually present; we can reject the null hypothesis of no trend. We can
then use this as a benchmark to see if subsamples of the time series
also indicate a significant increase.

@@

We then perform a subsampling routine to estimate the minimum time
required \(T_{min}\) (similar to Gerber, DeMaster, and Kareiva (1999)).
This is the same routine we used for the results in the main manuscript.

\begin{enumerate}
  \item We first extract all contiguous subsamples of the time series. This leads to 34 two-year subsamples, 33 two-year subsamples, and so forth until a single 35-year subsample.  
  \item For each subsample, we conduct linear regression and extract model coefficients and p-values.    \item We can call each set of subsamples, of the same length, a set. The fraction of subsamples within a set that show significant trends (significant slope coefficient $\alpha$ less than 0.05) is the statistical power. It is important to note that we only consider subsamples to be significant if they are significant in the same direction as the complete 35-year time series. In other words, we are conducting a one-tailed test.
  \item We can then plot statistical power as a function of time series length (Fig. \ref{fig:empirical_approach_example}b). As expected, we can see that power increases with the more years that are sampled.
  \item  Then, we determine an appropriate level of statistical power that we find acceptable. Traditionally, this has been at 0.8, however, this is purely historical. Statistical power of 0.8 implies if a true trend is present, or there is a real change in abundance, then we will detect this trend 0.8 fraction of the time. 
  \item We then determine the minimum time series length ($T_{min}$) required to achieve that level of statistical power. Here, $T_{min}$ is the first point in Fig. \ref{fig:empirical_approach_example}b where following points are also above 0.8. In this example, $T_{min}$ is 22. 
\end{enumerate}

Therefore, a minimum of 22 years of continuous monitoring are required
(for 0.8 statistical power at 0.05 significance level) to determine
long-term changes in abundance.

@@

\subsection{Predictors of minimum time
required}\label{predictors-of-minimum-time-required}

@@

In the main text, we explained how the minimum time required strongly
correlated with the trend strength, temporal autocorrelation, and
variance in population size. Here, we use a generalized linear model
framework with a Poisson error structure to determine explanatory
variables of the minimum time required. We use the same 878 populations
as in the main text. In figure \ref{fig:poisson_model}, we show a set of
residual plots for the regression. We then show the coefficient
estimates and levels of significance in table \ref{table:model_output}.

@@

In the main manuscript, we examined the minimum time required to detect
a significant trend in abundance over time using linear regression. As
detailed in the main manuscript the minimum time required was around 15,
but there was a wide distribution. Therefore, we were interested in
potential explanatory variables of the minimum time required. In the
main manuscript, we examined characteristics of the time series itself,
like variability, autocorrelation, and the trend in abundance over time.
Here, we combined our time series data with a data on life history
characteristics of amniotes from Myhrvold et al. (2015). There was life
history information available for 547 populations representing 315
different species, all of which were birds (Aves class).

@@

We then correlated minimum time required for each population with its
corresponding life history characteristics. In figure
\ref{fig:biological_correlates} we examined minimum time required versus
generation length (years), litter size (n), adult body mass (grams),
maximum longevity (years), egg mass (grams), and incubation (days). None
of these variables had much explanatory power in accounting for the
variance in the minimum time required. We ran a generalized linear model
(with a Poisson error structure) and found that these six variables only
accounted for 6.87\% of the explained deviance in minimum time required.

@@

Estimates of \(T_{min}\) depend strongly on the values used for the
statistical significance level (\(\alpha\)) and the probability of type
II error (\(\beta\)), both of which are set by the practitioner. Again,
statistical power is \(1-\beta\). Here we used simulations of the model
described in the main manuscript. The model simulates linear trends in
population abundance. We explored how estimates of \(T_{min}\) are
affected by changes in each of these parameters. We see that the minimum
time required increases with increases in statistical power or decreases
with increases in the threshold for statistical significance (Fig
\ref{fig:min_time_vs_alpha_beta}).

@@

In the main text, we evaluated the minimum time required to determine
long-term trends in abundance via linear regression. This process
examined linear trends in abundance over time. Here we examine the
minimum time required to estimate long-term trends that are either
exponential growth or decay. We use the same methods as described in the
previous sections, but we take the \(log\) of population density, or
abundance.

@@

We see that the distribution of \(T_{min}\) is almost identical to that
in the main manuscript (Fig. \ref{fig:min_time_dist_log_pop}). This is
perhaps not surprising as most time series that would significantly
increase or decrease linearly would probably also significantly increase
or decrease at an exponential rate. Further, the calculations here and
in the main manuscript both use linear regression. Therefore both
calculations estimate the same number of parameters.

@@

In the main text, we showed how a simple population model could be
simulated repeatedly to estimate the power obtained with time series of
increasing length (Bolker 2008; Johnson et al. 2015). The model in the
main text simulated linear population growth with only a slope
coefficient, y-intercept, and noise parameter required. This model is
purely phenomenological and does not include any species life history.
Here, we use the same routine as the main text, but simulate from a more
biologically-realistic population model. We use the model described in
White, Nagy, and Gruber (2014). The model is a stochastic,
age-structured population model that includes density-dependence for
lemon sharks (\emph{Negaprion brevirostris}) in Bimini, Bahamas.

@@

Here, we parameterize the model for a situation where adult mortality
rate is high enough to cause a population decline. In the same way we
simulated models in the main text, we simulate this more
biologically-realistic model for different lengths of time. For each
length of time, we calculate the statistical power. Similar to results
with the simpler model, statistical power generally increases with
longer sampling time (Fig. \ref{fig:shark_example}). In this example,
the minimum time required (\(T_{min}\)) t0 obtain at least 0.8
statistical power, given a significance level of 0.05, is 27 years.

@@

We applied the same calculations to 1032 population time series. We then
obtain a distribution of the minimum time required to measure the
``true" long-term growth rate (Fig. \ref{fig:min_time_growth_dist}). We
see a bimodal distribution with many populations required 30+ years to
estimate the long-term growth rate. The large number of short years
required is due to cases where the entire time series is consistently
increasing or decreasing at the same rate.

@@

In the main text, we examined the minimum time required to identify a
trend in abundance via linear regression. This approach allowed us to
identify increases or decreases, but a linear model may not always be a
good fit. Generalized additive models (GAMs) are more general than
general linear models and allow more flexibility (Wood 2006). GAMs are
models where a response variable depends on unknown smooth functions of
explanatory variables. GAMs, therefore, can identify relationships
between response variables and explanatory variables that are non-linear
and perhaps more complicated. The downside of GAMs is they typically
require more data and are also prone to overfitting.

Here, we conduct the same analyses in the main text, but instead
calculate the minimum time series required to detect trends over time
according to a GAM model. We hypothesized that GAMs should require less
time to detect a trend as they are more flexible than linear regression.
We provide an example that shows statistical power increases with more
time sampled (Fig. \ref{fig:gam_example}).

@@

\hypertarget{refs}{}
\hypertarget{ref-Bolker2008}{}
Bolker, Benjamin M. 2008. \emph{Ecological Models and Data in R}. 1st
ed. Princeton, New Jersey: Princeton University Press.

\hypertarget{ref-Gerber1999}{}
Gerber, L R, D P DeMaster, and P M Kareiva. 1999. ``Gray whales and the
value of monitering data in implementing the U.S. endangered species
act.'' \emph{Conservation Biology} 13 (5): 1215--9.

\hypertarget{ref-Johnson2015}{}
Johnson, Paul CD, Sarah JE Barry, Heather M Ferguson, and Pie Müller.
2015. ``Power analysis for generalized linear mixed models in ecology
and evolution.'' \emph{Methods in Ecology and Evolution} 6 (2): 133--42.
doi:\href{https://doi.org/10.1111/2041-210X.12306}{10.1111/2041-210X.12306}.

\hypertarget{ref-Myhrvold2015}{}
Myhrvold, Nathan P., Elita Baldridge, Benjamin Chan, Dhileep Sivam,
Daniel L. Freeman, and S.K. Morgan Ernest. 2015. ``An amniote
life-history database to perform comparative analyses with birds,
mammals, and reptiles.'' \emph{Ecology} 96 (11): 3109.

@@

\hypertarget{ref-RCoreTeam2016}{}
R Core Team. 2016. ``R: A language and environment for statistical
computing.'' Vienna, Austria: R Foundation for Statistical Computing.
\url{https://www.r-project.org/}.

@@

\hypertarget{ref-Wood2006}{}
Wood, S.N. 2006. \emph{Generalized Additive Models: An Introduction with
R}. Chapman Hall/CRC.
